Automatic discovery of cell types and microcircuitry from neural connectomics

### 2015-02-23 docs/paper.tex
added cluster comparison, cleaned up plots

\documentclass{article}
\usepackage[american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage{fixme}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
%\usepackage{algorithm}
%\usepackage{algorithmic}
\usepackage{amsmath}
%\usepackage{subcaption}
%\usepackage{multicol}
\usepackage{csquotes}
\usepackage{pdfpages}
\usepackage{placeins}

@@

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{fig1b.ai}}
  \caption{Deriving circuitry and types from connectomics data a. As
    input we take the connectivity between cells (b), the distance
    between them (c), the depth of the cell bodies (d), and the depth
    profile of the synapses (e). f. Our algorithm discovers hidden
    cell types in this connectivity data by assuming all cells of a
    type share a distance-dependent connectivity profile, similar
    depth, and a similar synaptic density profile, with cells of other
    types.  This results in a clustering of the cells by those hidden
    types. f.) shows the cell connectivity matrix with cells of the
    same type grouped together. g.) shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. h. shows the probability of connection between two cell
    types that very rarely connect -- there's a background ``base''
    connection rate to account for errors in data, but the probability
    is very low.  We can plot this these types (i) to show the
    relationship and spatial connections, and how probable various
    types are to connect to one another. j.) The connectivity between
    our discovered types changes as a function of distance. }
\label{fig:overview}
\end{figure}

@@

\subsection{Metrics}
To evaluate the quality of the model fit, we need to use information that quantifies aspects of the data for which we have ground truth information. We focus on two aspects of performance. First, if the model works well, then the probability that a pair of neurons is of the same type should be high if the neurons actually are of the same type. Second, the model should assign a high probability of connection between two cells if they have a connection in the underlying data. We term these two factors clustering accuracy and link prediction accuracy.

@@

\subsection{Validation with simulated data where ground truth is known}
To validate our model, we performed a series of simulations to test if
the model can accurately recover the true underlying network structure
and cell type identity.  We thus simulate data for which we know the
correct structure and compare the estimated structure based on the
algorithm (see methods) with the one we used for simulation. We find
that the model does a good job of recovering the correct number of
cell types, (fig~\ref{fig:synthetic}a), the the cell identities
(fig~\ref{fig:synthetic}b), and the spatial extent of each type
(fig~\ref{fig:synthetic}c).  For comparison, we show the results from instead using the infinite
stochastic block model (fig~\ref{fig:synthetic}a, b, c, black line) which assumes that only cell type matters, and thus finds
small neighborhoods of connected nodes (instead of global connectivity
patterns). This contrast shows that while the regular block model can not correctly deal with distance dependent connectivity, our model can. Our model converges relatively quickly (see supplemental) to an estimate of
the most probable values for the cell types, which is enabled by using
a combination of simulated annealing and parallelized Markov chain
Monte Carlo (see methods for details). Thus our model at least is promising to be applied to biological datasets.

\subsection{Model Mismatch}

@@

\subsection{Sensitivity to Edge Effects}

\begin{figure}
  \centering 
  % generated from experiments/edgeeffects/edgeeffects.ipynb
  \centerline{\includegraphics[width=3.5in]{edgeeffects.pdf}}
  \caption{edge effects}
\label{fig:synthetic}
\end{figure}

\fxwarning{cut down on the number of ticks, this will be a small figure, homogenize fonts}

Connectomic efforts so far have reconstructed only small sections of
tissue at a time. Consequently, many connections to cells outside that
tissue volume will be lost. We are concerned that this selective
elimination of connectivity along the boundary might give the
appearance of distance-dependent connectivity when there is none. We
thus performed simulations to check if edge effects could destroy
spatial structure and if edge effects could introduce artificial,
spurious spatial structure. We measure the degree to which the model
infers distance dependent connectivity for two synthetic datasets --
one with distance-dependence and one without. We find that, if there
is no distance-dependence, edge effects do not artificially introduce
distance dependence. However, if the section we are examining is too
small, our model can miss the distance dependence. Thus with respect
to distance-dependent connectivity inference, our model errs on the
side of caution. But we also find that for spatial extend that is
similar to the currently available datasets the effects of this are
quite limited.

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics{mouseretina.ai}}
  \caption{Discovering cell classes in the mouse retina connectome.
    a.) Input connectivity data for 950 cells for which soma positions
    were known. b. clustered connectivity matrix. c.) connectivity
    diagram showing our clusters, as well as the cell depth and
    anatomist-labeled cell type of each cell. Our model shows moderate
    agreement with the cell types traditionally identified by
    anatomist laminar-specific connectivity patterns. 
    d.) Connectivity between our clusters as a function
    of distance -- the cluster consisting primarily of retinal
    ganglion cells (lower left-center) exhibits the expected near and
    far connectivity. e.)  The spatial distribution of our cell types
    -- each cell type tesselates space.}
\label{fig:mouseretina}
\end{figure}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics{comparison.ai}}
  \caption{\textbf{Visualizing type inference uncertainty.} Our fully Bayesian model gives a confidence estimate (posterior probability) that any two given cells are of the same type. In b.) we visualize that cell-cell coassignment matrix, showing the probability that cell i is of the same type as cell j on a range from 0.0 to 1.0. The block structure shows subsets of cells which are believed to all belong to the same type. For comparison, a.) shows the anatomist-defined type for each cell, grouped broadly into the coarse types identified in the previous figure. \textbf{Link vs cluster accuracy} c. A comparison of the
    predictive accuracy (area under the curve) for hand-labeled
    anatomical data, versus inclusion of additional sources of
    information, as well as the clustering accuracy. Note that our model sacrifices very little predictive accuracy for additional clustering accuracy. By comparison, conventional methods fail at one or both.  d. ) The spatial extent (in depth and area) of the
    types identified by humans and our various algorithmic
    approaches.}

\label{fig:mouseretina_compare}
\end{figure}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=4.7in]{celegans2.ai}}
  \caption{Discovering connectivity and type in
    \textit{C. elegans}. Discovered types a.) The cell/cell
    coassignment matrix identifies the posterior probability of any
    two cells in \textit{c. elegans} being of the same type. The large
    visible clusters roughly correspond to a more-coarse version of
    types identified by neuroanatomists seen in b.) In particular,
    robust subpopulations are found corresponding to divisions of
    moter, interneuron, and sensory neurons. c.) The location of the
    cell soma, revealing how spatial localizatinion correlates with
    type. d.) The ROC curves for held-out link probability for both
    the electrical synapses (gap junctions) and chemical synapses in
    c. elegans. e.) The posterior distribution of the area under the
    ROC curve for the curves in d. f.) Measurements of the
    agreement of our identified cell types compared to neuroanatomists. 
    See methods for an explanation of metrics. }
  \label{fig:celegans}
\end{figure}

@@

\fxwarning{Add chip back in}

@@

\section*{Measuring clustering similarity}

The adjusted rand index (ARI) is a measure of the similarity of two
different clusterings \autocite{Hubert1985} -- two identical clusters
have an ARI of 1.0 while progressively more dissimilar clusters have
lower ARIs, becoming negative as the clustering gets anti-correlated.

>>>>

\documentclass{article}
\usepackage[american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage[margin=true,inline=true,index=true, status=draft]{fixme}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
%\usepackage{algorithm}
%\usepackage{algorithmic}
\usepackage{amsmath}
%\usepackage{subcaption}
%\usepackage{multicol}
\usepackage{csquotes}
\usepackage{pdfpages}
\usepackage{placeins}

@@

\listoffixmes

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=110mm]{fig1b.ai}}
  \caption{Deriving circuitry and cell types from connectomics data
    a. As input we take the connectivity between cells (b), the
    distance between them (c), the depth of the cell bodies (d), and
    the depth profile of the synapses (e). f. Our algorithm discovers
    hidden cell types in this connectivity data by assuming all cells
    of a type share a distance-dependent connectivity profile, similar
    depth, and a similar synaptic density profile, with cells of other
    types.  This results in a clustering of the cells by those hidden
    types. f.) shows the cell connectivity matrix with cells of the
    same type grouped together. g.) shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. h. shows the probability of connection between two cell
    types that very rarely connect -- there's a background ``base''
    connection rate to account for errors in data, but the probability
    is very low.  i.) shows that we also recover the expected
    laminarity of types and the depth-specific synaptic
    connectivity. j.) We then plot how the connectivity between these
    types changes as a function of distance between the cell bodies to
    better understand short-range and long-range connectivity
    patterns. }

\label{fig:overview}
\end{figure}

@@

\subsection*{Metrics}
To evaluate the quality of the model fit, we need to use information that quantifies aspects of the data for which we have ground truth information. We focus on two aspects of performance. First, if the model works well, then the probability that a pair of neurons is of the same type should be high if the neurons actually are of the same type. Second, the model should assign a high probability of connection between two cells if they have a connection in the underlying data. We term these two factors clustering accuracy and link prediction accuracy.

@@

\subsection*{Validation with simulated data where ground truth is known}
To validate our model, we performed a series of simulations to test if
the model can accurately recover the true underlying network structure
and cell type identity.  We thus simulate data for which we know the
correct structure and compare the estimated structure based on the
algorithm (see methods) with the one we used for simulation. We find
that the model does a good job of recovering the correct number of
cell types, (fig~\ref{fig:synthetic}a), the the cell identities
(fig~\ref{fig:synthetic}b), and the spatial extent of each type
(fig~\ref{fig:synthetic}c).  For comparison, we show the results from instead using the infinite
stochastic block model (fig~\ref{fig:synthetic}a, b, c, black line) which assumes that only cell type matters, and thus finds
small neighborhoods of connected nodes (instead of global connectivity
patterns). This contrast shows that while the regular block model can not correctly deal with distance dependent connectivity, our model can. Our model converges relatively quickly (see supplemental) to an estimate of
the most probable values for the cell types, which is enabled by using
a combination of simulated annealing and parallelized Markov chain
Monte Carlo (see methods for details). Thus our model at least is promising to be applied to biological datasets.

\subsection*{Model Mismatch}

@@

\subsection*{Sensitivity to Edge Effects}

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=3.5in]{edgeeffects.pdf}}
  \caption{We generate two sets of synethetic data, one with
    spatially-dependent connectivity and one without. We measure the
    variance in the connectivity-distance plot for randomly selected
    regions of each dataset, ranging from single-cells to the entire
    volume. We see that while selecting too small of a region can 
    destroy the appearance of distance-dependent connectivity, 
    it does not create it in non-spatial data}
\label{fig:edgeeffects}
\end{figure}

Connectomic efforts so far have reconstructed only small sections of
tissue. Consequently, many connections to cells outside that tissue
volume will be lost. We are concerned that this selective elimination
of connectivity along the boundary might give the appearance of
distance-dependent connectivity when there is none. We thus performed
simulations to check if edge effects could destroy spatial structure
and if edge effects could introduce artificial, spurious spatial
structure. We measure the degree to which distance-dependent effects
can arise from selecting regions that are smaller than the ``scale''
of connectivity (fig ~\ref{fig:edgeeffects}). We do this by generating
two collections of synthetic datasets -- one with distance-dependent
connectivity and one without.  We then in each dataset randomly
examine contiguous circular regions with area varying from zero to the
entire volume, and empirically calculate the spatial variance in
type-dependent connectivity.  We find that, if there is no
distance-dependence, edge effects do not artificially introduce
distance dependence. However, if the section we are examining is too
small, our model can miss the distance dependence. Thus with respect
to distance-dependent connectivity inference, our model errs on the
side of caution. But we also find that for spatial extend that is
similar to the currently available datasets the effects of this are
quite limited.

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=160mm]{mouseretina2.ai}}
  \caption{Discovering cell classes in the mouse retina connectome. Here
we show the maxium a posteriori (MAP) estimate for the types in the mouse
retina data. 
    a.) Input connectivity data for 950 cells for which soma positions
    were known. b. clustered connectivity matrix, each arbitrary color
    corresponds to a single type, and will be used to identify that type in the remainder
    of the plot. c.)  The spatial distribution of our cell types
    -- each cell type tesselates space. Colors correspond to those in b.)
    d.) Connectivity between our clusters as a function
    of distance -- the cluster consisting primarily of retinal
    ganglion cells (brown node on graph) exhibits the expected near and
    far connectivity.}
\label{fig:mouseretina}
\end{figure}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics{comparison.ai}}
  \caption{\textbf{Visualizing type inference uncertainty.} Our fully
    Bayesian model gives a confidence estimate (posterior probability)
    that any two given cells are of the same type. In a.) we visualize
    that cell-cell coassignment matrix, showing the probability that
    cell i is of the same type as cell j on a range from 0.0 to
    1.0. The block structure shows subsets of cells which are believed
    to all belong to the same type. For comparison, b.) shows the
    anatomist-defined type for each cell, grouped broadly into the
    coarse types identified in the previous figure. c.) for each cell
    in the matrix, c.) shows the soma depth as well as the depth
    histogram for all putative synaptic contacts. \textbf{Link vs
      cluster accuracy} d.) The posterior distribution of ROC curves
    from 10-fold cross validation when predicting connectivity, as
    well as e.) the area under the curve and f.) the type agreements
    with known neuroanatomist types. g. A comparison of the predictive
    accuracy (area under the curve) for hand-labeled anatomical data,
    versus inclusion of additional sources of information, as well as
    the clustering accuracy. Note that our model sacrifices very
    little predictive accuracy for additional clustering accuracy. By
    comparison, conventional methods fail at one or both.  d. ) The
    spatial extent (in depth and area) of the types identified by
    humans and our various algorithmic approaches.}
\label{fig:mouseretina_compare}
\end{figure}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=7.0in]{celegans2.ai}}
  \caption{Discovering connectivity and type in
    \textit{C. elegans}. a.) posterior distribution on cell
    connectivity as a function of discovered type, similar to
    figure~\ref{mouseretina}. In b.) we plot neuroanatomist-derived
    types along with their labels. Our model shows high probability of
    motor neurons, sensory neurons, and various interneuron classes
    being of the same type. Soma positions along the body axis are
    plotted in c.) where we see that we cluster spatially-distributed
    motor neurons together, whereas head sensory neurons are more
    likely to be grouped together as well.  d.) The ROC curves for
    held-out link probability for both the electrical synapses (gap
    junctions) and chemical synapses in c. elegans. e.) The posterior
    distribution of the area under the ROC curve for the curves in
    d. f.) Measurements of the agreement of our identified cell types
    compared to neuroanatomists. The high compleneness but low
    homogeneity (and corresponding low ARI) reflects our model's
    tendency to group multiple types into a single type.}
  \label{fig:celegans}
\end{figure}

@@

\subsection*{Types and connectivity in artificial structures}
\begin{figure}
  \centering 
  \centerline{\includegraphics[width=4in]{mos6502.ai}}
  \caption{Discovering connectivity and type the 6502 microprocessor.
    a.) is the micrograph of the original microprocessor, with the
    region containing the registers under study highlighted. b.) Our
    graph consists of the interconnections of MOS field-effect
    transistors with three terminals, Gate, C1, and C2. Reconstruction
    technique did not permit resolution of C1 and C2 into source and
    drain. c.) The spatial distribution of the transistors in each
    cluster show a clear pattern d.) The clusters and connectivity
    versus distance for connections between Gate and C1, Gate and C2,
    and C1 and C2 terminals on a transistor. Red and teal types have a
    terminal pulled down to ground and mostly function as
    inverters. Purple class are clocked, stateful transistors, orange
    control the ALU and yellow control the special data bus (SDB).}
  \label{fig:mos6502}
\end{figure}

To show the applicability of our method to other connectome-style
datasets, we obtained the spatial location and interconnectivity of
the transistors in a classic microprocessor, the MOS Technology 6502
(used in the Apple II) \autocite{James2010}. Computer architects use
common patterns of transistors when designing circuits, with each
transistor having a ``type'' in the circuit. We identified a region of
the processor with complex but known structure containing the primary
8-bit registers X, Y, and S (fig~\ref{fig:mos6502}).

Our algorithm identifies areas of spatial homogeneity that mirror the
known structure in the underlying architectural circuit, segmenting
transistor types recognizable to computer architects. Using the
original schematics, we see that one identified type contains the
``clocked'' transistors, which retain digital state. Two other types
contain transistors with pins C1 or C2 connected to ground, mostly
serving as inverters.  An additional identified type controls the
behavior of the three registers of interest (X, Y, and S) with respect
to the SB data bus, either allowing them to latch or drive data from
the bus. The repeat patterns of spatial connectivity are visible in
figure~\ref{fig:mos6502}c, showing the man-made horizontal and
vertical layout of the same types of transistors.

\subsection*{Microprocessor}
We extracted the connection graph for the transistors in the MOS6502
\autocite{visual6502source}. Each transistor has three terminals (gate,
source, drain), but the methods of the original dataset were unable to
consistently resolve which of the C1 and C2 terminals were source and
drain, leading to ambiguity in our encoding. We identified a region
consisting of three registers X, Y, and S via visual
inspection and focused our efforts there. We created a total of six
connectivity matrices by examining possible terminal pairings. One
graph, for example, $R^{gc_1}(e_i, e_j)=1$ if transistor $e_j$ and
$e_j$ are connected via pins $g$ and $c_1$.

@@

\section*{Measuring clustering similarity}
\fxwarning{Add an explanation of ARI, etc.} 
The adjusted rand index (ARI) is a measure of the similarity of two
different clusterings \autocite{Hubert1985} -- two identical clusters
have an ARI of 1.0 while progressively more dissimilar clusters have
lower ARIs, becoming negative as the clustering gets anti-correlated.
